---
title: "Visualizing TPC Data (1999 vs. 2024)"
author: "Taylor Hatcher"
date: '2025-01-28'
output:
  pdf_document: default
  html_document: default
---


```{r setup, include=FALSE}
# Load required libraries
library(ggplot2)
library(data.table)
library(dplyr)
library(patchwork)
library(reshape2)
library(viridis)
library(tidyverse)
library(lubridate)
```
```{r}
# Load Data 
getwd()
tpcvis <- read.csv("PastPresentFilteredConstantTpc2024.csv")
```
```{r}
# Filter Data 
tpcvis_current <- tpcvis %>%
filter(time.per == "current")

# Log Scale Growth Rate Calculation
current_plot_log <- ggplot(tpcvis_current, aes(x = temp, y = rgrlog, color = mom)) +
  geom_point() +
  facet_grid(durbin ~ instar) +
  theme_bw() +
  xlab("Temperature (°C)") +
  ylab("RGR (mg/mg/h)") +
  ggtitle("Present Data (2024): Log Scale") +
  ylim(0, 0.20) +
  scale_color_viridis(discrete = TRUE)
current_plot_log
```

``` {r}
# Arithmetic Scale Growth Rate Calculation 
current_plot_arith <- ggplot(tpcvis_current, aes(x = temp, y = rgrarith, color = mom)) +
  geom_point() +
  facet_grid(durbin ~ instar) +
  theme_bw() +
  xlab("Temperature (°C)") +
  ylab("Arithmetic Growth Rate") +
  ggtitle("Present Data (2024): Arithmetic Scale") +
  ylim(0, 12.00) +
  scale_color_viridis(discrete = TRUE)
current_plot_arith
```


```{r}
# Combined Scale Comparison (Current Data)
combined_current_scale_comparison_plot <- current_plot_log + current_plot_arith + plot_layout(ncol = 1)
combined_current_scale_comparison_plot
```

# Historic Data Analysis
## Filter Historic Data
``` {r}
tpcvis_past <- tpcvis %>%
  filter(time.per == "past")
```
## Log Scale Growth Rate
``` {r}
historic_plot_log <- ggplot(tpcvis_past, aes(x = temp, y = rgrlog, color = mom)) +
  geom_point() +
  facet_grid(durbin ~ instar) +
  theme_bw() +
  xlab("Temperature (°C)") +
  ylab("RGR (mg/mg/h)") +
  ggtitle("Historic Data (1999): Log Scale") +
  ylim(0, 0.20) +
  scale_color_viridis(discrete = TRUE)
historic_plot_log
```

## Arithmetic Scale Growth Rate
```{r}
historic_plot_arith <- ggplot(tpcvis_past, aes(x = temp, y = rgrarith, color = mom)) +
  geom_point() +
  facet_grid(durbin ~ instar) +
  theme_bw() +
  xlab("Temperature (°C)") +
  ylab("Arithmetic Growth Rate") +
  ggtitle("Historic Data (1999): Arithmetic Scale") +
  ylim(0, 12.00) +
  scale_color_viridis(discrete = TRUE)
historic_plot_arith
```


## Combined Scale Comparison (Historic Data)
``` {r}
combined_historic_scale_comparison_plot <- historic_plot_log + historic_plot_arith + plot_layout(ncol = 1)
combined_historic_scale_comparison_plot
```

# Comparison of 1999 vs. 2024 Data

## Logarithmic Scale Comparison
```{r}
tpc.logplot <- ggplot(tpcvis, aes(x = temp, y = rgrlog, color = time.per)) +
  geom_point(alpha = 0.5) +
  facet_grid(durbin ~ instar) +
  theme_bw() +
  xlab("Temperature (°C)") +
  ylab("RGR (mg/mg/h)") +
  ggtitle("1999 vs. 2024: Logarithmic Scale RGR") +
  ylim(-0.02, 0.12) +
  scale_color_manual(values = c("current" = "#EE6A50", "past" = "#7AC5CD"))
tpc.logplot
```

## Arithmetic Scale Comparison
```{r}
tpc.arithplot <- ggplot(tpcvis, aes(x = temp, y = rgrarith, color = time.per)) +
  geom_point(alpha = 0.5) +
  facet_grid(durbin ~ instar) +
  theme_bw() +
  xlab("Temperature (°C)") +
  ylab("Arithmetic RGR (mg/mg/h)") +
  ggtitle("1999 vs. 2024: Arithmetic Scale RGR") +
  ylim(-0.02, 11.0) +
  scale_color_manual(values = c("current" = "#EE6A50", "past" = "#7AC5CD"))
tpc.arithplot
```


# Aggregated Data Analysis
## Logarithmic Scale Aggregation
```{r}
tpc.agglog <- tpcvis %>%
  group_by(temp, time.per, durbin, instar) %>%
  summarise(
    mean = mean(rgrlog, na.rm = TRUE),
    se = sd(rgrlog, na.rm = TRUE) / sqrt(n())
  )
tpc.plotlog <- ggplot(tpc.agglog[tpc.agglog$durbin %in% c(6, 24), ], aes(x = temp, y = mean, color = time.per, group = time.per)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.3, alpha = 0.6) +
  facet_grid(durbin ~ instar) +
  theme_bw() +
  xlab("Temperature (°C)") +
  ylab("RGR (mg/mg/h)") +
  ggtitle("Aggregated Logarithmic Scale Data") +
  ylim(-0.10, 0.14) +
  scale_color_manual(values = c("current" = "#EE6A50", "past" = "#7AC5CD"))
tpc.plotlog
```

## Arithmetic Scale Aggregation
```{r}
tpc.aggarith <- tpcvis %>%
  group_by(temp, time.per, durbin, instar) %>%
  summarise(
    mean = mean(rgrarith, na.rm = TRUE),
    se = sd(rgrarith, na.rm = TRUE) / sqrt(n())
  )
tpc.plot.arith <- ggplot(tpc.aggarith[tpc.aggarith$durbin %in% c(6, 24), ], aes(x = temp, y = mean, color = time.per, group = time.per)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.3, alpha = 0.6) +
  facet_grid(durbin ~ instar) +
  theme_bw() +
  xlab("Temperature (°C)") +
  ylab("Arithmetic RGR (mg/mg/h)") +
  ggtitle("Aggregated Arithmetic Scale Data") +
  ylim(-0.10, 6.00) +
  scale_color_manual(values = c("current" = "#EE6A50", "past" = "#7AC5CD"))
tpc.plot.arith
```

```{r}

# Filter na's from rgrlog out
tpcvis_cleancurrent <- tpcvis_current %>% filter(is.finite(rgrlog))

# Temp is a factor for analyzing
tpcvis_cleancurrent$temp <- as.factor(tpcvis_cleancurrent$temp)
```
## Analysis approach
# 1. Fit the mixed model with a quadratic time effect
# 2. Check model fit using summary & ANOVA
# 3. Simulate residuals with DHARMa
# 4. Run residual diagnostics 
# 5. Compare models (linear vs. quadratic)
# 6. Refit the final model using REML
# 7. Visualize residuals and predictions using DHARMa 
```{r}
library(lme4)
```
# Fit mixed model with a quadratic time effect--
#### rgrlog= response variable, poly(duration, 2) - models duration as a quadratic polynomial, temp is a catergorical fixed effect, (1 | ID) random intercept for indivdual

```{r}
mod.lmer <- lmer(rgrlog ~ poly(duration, 2) + temp + (1 | ID),
                 data = tpcvis_cleancurrent,
                 REML = FALSE)

summary(mod.lmer)
anova(mod.lmer)
```
# Simulate residuals
```{r}
sim_res <- simulateResiduals(fittedModel = mod.lmer)

print(sim_res)


plot(sim_res)
```
# Check for uniformity
```{r}
testUniformity(sim_res)
```
# Check for overdispersion
```{r}
testDispersion(sim_res)
```
# Check for outliers
```{r}
testOutliers(sim_res)

mod.lmer_linear <- lmer(rgrlog ~ duration + temp + (1 | ID), 
                        data = tpcvis_cleancurrent, REML = FALSE)

anova(mod.lmer_linear, mod.lmer)  # Compare linear vs. quadratic



mod.lmer_final <- lmer(rgrlog ~ poly(duration, 2) + temp + (1 | ID), 
                       data = tpcvis_cleancurrent, 
                       REML = TRUE)  # Now using REML

```
```{r}
library(DHARMa)
```
# Simulate residuals
```{r}
sim_res <- simulateResiduals(fittedModel = mod.lmer_final)

# Plot residual diagnostics
plot(sim_res)

```
```{r}
library(ggplot2)
```
# Extract predictions
```{r}
tpcvis_cleancurrent$predicted <- predict(mod.lmer_final)

ggplot(tpcvis_cleancurrent, aes(x = predicted, y = rgrlog)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", col = "red") +
  labs(title = "Observed vs. Predicted Values",
       x = "Predicted Log Relative Growth Rate",
       y = "Observed Log Relative Growth Rate") +
  theme_minimal()
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.



